---
title: "Cyclistic Data Cleaning Changelog"
author: "James Kenneth Ines"
date: "2025-07-06"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
---



# Changelog
This file contains the notable changes to the project

## Version 0 (07-06-2025)
This version was the original version that was pre-cleaned to some extent using Google Sheets and was then ingested in Rstudio. The changes, fixes, and additions were all done in Google Sheets.

### Fixes
- No duplicates found for both Copy of CyclisticTrips2019_q1_20250706_v01 and Copy of CyclisticTrips2020_q1_20250706_v01
- No whitespaces needing to be trimmed found

### Changes
- Cleared all formatting

### New
- Made a copy in Google Sheets

### Removed
- (Trips2020_q1)Deleted row with ride_id 157EAA4C4A3C8D36


## Version 1 (07-07-2025)
Started using R to clean the data


### New
- Set up environment

```{r}
getwd()
setwd("~/Desktop/Cyclistic Case Study 2025")
```

- Installed and loaded necessary cleaning packages

```{r message=FALSE, warning=FALSE}
# Install packages
install.packages("here", repos = "http://cran.us.r-project.org")
install.packages("skimr", repos = "http://cran.us.r-project.org")
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("tidyverse", repos = "http://cran.us.r-project.org")  # Includes dplyr, lubridate, and more

# Load packages
library(here)
library(skimr)
library(janitor)
library(tidyverse)  # Loads dplyr, lubridate, ggplot2, etc.
```
- Ingested datasets
```{r}
# Ingest data
q1_2019 <- read.csv("Copy of CyclisticTrips2019_q1_20250706_v01 - Trips2019_q1.csv")
q1_2020 <- read.csv("Copy of CyclisticTrips2020_q1_20250706_v01 - Trips2020_q1.csv")

```


- Summarized datasets

```{r}
skim_without_charts(q1_2019)
glimpse(q1_2019)
head(q1_2019)
colnames(q1_2019)

skim_without_charts(q1_2020)
glimpse(q1_2020)
head(q1_2020)
colnames(q1_2020)
```

### Removed
- (Trips2019_q1)Removed outliers in tripduration column that is less than 60 seconds and more than 86400 seconds(24 hours)

```{r warning=FALSE}
q1_2019_cleaned_v1 <- q1_2019 %>%
  filter(tripduration >= 60 & tripduration <=86400)

# Count rows before and after
cat("Original rows:", nrow(q1_2019), "\n")
cat("Cleaned rows:", nrow(q1_2019_cleaned_v1), "\n")
```

## Version 2 (07-08-2025)


### New
- (q1_2019 and q1_2020)Added weekday column
- (q1_2020) added tripduration column

```{r}
# Add weekday column for v2

install.packages("hms", repos = "http://cran.us.r-project.org")
library(hms)

q1_2019_cleaned_v2 <- q1_2019_cleaned_v1 %>%
  separate(start_time, into=c("start_date", "start_time"), sep=" ") %>% 
  separate(end_time, into=c("end_date", "end_time"), sep=" ") %>% 
  
  mutate(start_time = hms::as_hms(start_time),
         end_time = hms::as_hms(end_time)) %>%
  mutate(start_date = as.Date(start_date, format = "%m/%d/%Y"),
         end_date = as.Date(end_date, format = "%m/%d/%Y")) %>% 
  
  mutate(weekday = factor(weekdays(start_date), 
                          levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))


head(q1_2019_cleaned_v2 %>% select(start_time, end_time, weekday, tripduration))
nrow(q1_2019_cleaned_v2)
summary(q1_2019_cleaned_v2$tripduration)

# Added tripduration and weekday columns for q1_2020_cleaned_v1

q1_2020_cleaned_v1 <- q1_2020 %>%
  separate(started_at, into=c("start_date", "start_time"), sep=" ") %>% 
  separate(ended_at, into=c("end_date", "end_time"), sep=" ") %>% 
  
  mutate(start_time = hms::as_hms(start_time),
         end_time = hms::as_hms(end_time)) %>%
  mutate(start_date = as.Date(start_date, format = "%m/%d/%Y"),
         end_date = as.Date(end_date, format = "%m/%d/%Y")) %>% 
  # Add tripduration column
  mutate(tripduration = end_time - start_time) %>% 
  # Add weekday column
  mutate(weekday = factor(weekdays(start_date), 
                          levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))


head(q1_2020_cleaned_v1 %>% select(start_time, end_time, weekday, tripduration))
arrange(q1_2020_cleaned_v1, desc(tripduration))
```

### Fixes
- (Trips2020_q1)Removed outliers for tripduration

```{r}

q1_2020_cleaned_v2 <- q1_2020_cleaned_v1 %>%
  filter(tripduration >= 60 & tripduration <=86400)

head(q1_2020_cleaned_v2)
nrow(q1_2020_cleaned_v1)
nrow(q1_2020_cleaned_v2)
```
## Version 3 (07-10-2025)


### Fixes
- (q1_2019 and q1_2020)Cleaned column names
- (q1_2019 and q1_2020)Checked for NA values

### New
- Added tripduration_minutes column

```{r}
q1_2019_cleaned_v3 <- q1_2019_cleaned_v2 %>% 
  mutate(tripduration_minutes = as.numeric(difftime(end_time, start_time, units = "mins")))

clean_names(q1_2019_cleaned_v3) # Cleaned column names
colSums(is.na(q1_2019_cleaned_v3)) # check for NA values
head(q1_2019_cleaned_v3)

q1_2020_cleaned_v3 <- q1_2020_cleaned_v2 %>% 
  mutate(tripduration_minutes = as.numeric(difftime(end_time, start_time, units = "mins")))

clean_names(q1_2019_cleaned_v3) # Cleaned column names
colSums(is.na(q1_2019_cleaned_v3)) # check for NA values
head(q1_2020_cleaned_v3)
```



## Version 4 (07-14-2025)


### Fixes
- (q1_2019 and q1_2020)Noticed some whitespace and removed it

### Changes
- (q1_2019 and q1_2020)Removed asterisks on the station name addresses
- (q1_2019 and q1_2020)Converted station names to title-case
- (q1_2020)Changed column names to match q1_2019 column names

```{r}
names(q1_2020_cleaned_v3)
unique(q1_2020_cleaned_v3$start_station_name)
unique(q1_2020_cleaned_v3$end_station_name)

q1_2020_cleaned_v3 %>%
  filter(grepl("\\*", start_station_name) | grepl("\\*", end_station_name)) %>%
  select(start_station_name, end_station_name) %>%
  head()

q1_2020_cleaned_v4 <- q1_2020_cleaned_v3 %>%
  # Clean station names: remove whitespace, asterisks, and convert to title case
  mutate(start_station_name = trimws(start_station_name),  # Remove leading/trailing whitespace
         end_station_name = trimws(end_station_name),
         start_station_name = gsub("\\s*\\(\\*\\)|\\s*\\(temp\\)", "", start_station_name),  # Remove asterisks
         end_station_name = gsub("\\s*\\(\\*\\)|\\s*\\(temp\\)", "", end_station_name),
         start_station_name = tools::toTitleCase(tolower(start_station_name)),  # Convert to title case
         end_station_name = tools::toTitleCase(tolower(end_station_name)))

q1_2020_cleaned_v4 %>%
  filter(grepl("\\*", start_station_name) | grepl("\\*", end_station_name)) %>%
  select(start_station_name, end_station_name) %>%
  head()

# changing column names for possible data aggregation
names(q1_2020_cleaned_v4) [1] <- "trip_id"
names(q1_2020_cleaned_v4) [2] <- "bike_id"
names(q1_2020_cleaned_v4) [7] <- "from_station_name"
names(q1_2020_cleaned_v4) [8] <- "from_station_id"
names(q1_2020_cleaned_v4) [9] <- "to_station_name"
names(q1_2020_cleaned_v4) [10] <- "to_station_id"
names(q1_2020_cleaned_v4) [15] <- "usertype"


colnames(q1_2020_cleaned_v4)
```




